---
title: "Bike Use Analysis"
output: html_document
---

## Introduction and background
Using Cyclistic’s historical trip data to analyze and identify trends

## Installing and loading common packages and libraries
Here we are installing the tidyverse and ggplot2 packages which are important and useful for cleaning and visualizing our data

```{r}
#install.packages('tidyverse')
library(tidyverse)
#install.packages("ggplot2")
library(ggplot2)
```

## Loading your CSV files
Here we are creating data frames from each CSV file for the past 12 months and putting them into said data frames

```{r}
Dec_2021 <-read.csv("202112-divvy-tripdata.csv")
Jan_2022 <- read.csv("202201-divvy-tripdata.csv")
Feb_2022 <- read.csv("202202-divvy-tripdata.csv")
Mar_2022 <- read.csv("202203-divvy-tripdata.csv")
Apr_2022 <- read.csv("202204-divvy-tripdata.csv")
May_2022 <- read.csv("202205-divvy-tripdata.csv")
Jun_2022 <- read.csv("202206-divvy-tripdata.csv")
Jul_2022 <- read.csv("202207-divvy-tripdata.csv")
Aug_2022 <- read.csv("202208-divvy-tripdata.csv")
Sep_2022 <- read.csv("202209-divvy-tripdata.csv")
Oct_2022 <- read.csv("202210-divvy-tripdata.csv")
Nov_2022 <- read.csv("202211-divvy-tripdata.csv")
```

##  Checking out the column names 
Here we are exploring a few key details about the data provided

```{r}
colnames(Dec_2021)
colnames(Jan_2022)
colnames(Feb_2022)
colnames(Mar_2022)
colnames(Apr_2022)
colnames(May_2022)
colnames(Jun_2022)
colnames(Jul_2022)
colnames(Aug_2022)
colnames(Sep_2022)
colnames(Oct_2022)
colnames(Nov_2022)
```

## Altering the column names
We are changing the column names so that they are easier to understand
```{r}
colnames(Dec_2021) <- c("trip_id", "bike_id","start_time"
                   , "end_time", "from_station_name","from_station_id", "to_station_name", "to_station_id", "from_lat"
                   , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Jan_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Feb_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Mar_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Apr_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(May_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Jun_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Jul_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Aug_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Sep_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Oct_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
colnames(Nov_2022) <- c("trip_id", "bike_id", "start_time"  
                        , "end_time", "from_station_name"
                        , "from_station_id", "to_station_name" 
                        , "to_station_id", "from_lat"
                        , "from_lng", "to_lat", "to_lng", "usertype")
```


## Checking out some basic details
Here we look at the summary for one of the parameters 
```{r}
summary(Dec_2021)
```

## Binding together
Here we bind all the data from the past 12 months into one table so that it is easier to analyze
```{r}
all_trips <- bind_rows(Dec_2021, Jan_2022, Feb_2022,
                       Mar_2022, Apr_2022, May_2022,
                       Jun_2022, Jul_2022, Aug_2022,
                       Sep_2022, Oct_2022, Nov_2022)
```

## Checking up on some details
Here we check up on a couple of details concerning the new table we have created all_trips
```{r}
head(all_trips)
tail(all_trips)
colnames(all_trips)
dim(all_trips)
str(all_trips)
summary(all_trips)
```

## Next we will add new, columns for dates, day, month, year, and ride length
```{r}
all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
```

## Removing unused columns
Removing the long and lat columns as the are not important for the analysis
```{r}
all_trips <- all_trips %>%  
  select(-c(from_lat, from_lng, to_lat, to_lng))
```

## Converting from one variable to another
Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(all_trips$ride_length) 
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

## Cleaning data
Here we will remove bad data from the dataset so that it won't affect our analysis
```{r}
all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR" | all_trips$ride_length<0),]
```

Now we check to see if the data is cleaned
```{r}
summary(all_trips_v2$ride_length)
```

## Now we look at the stats based in user type
Comparing the summary stats of casual and annual members
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = min)
```

Checking the average riode time for annual and casual members
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
```

We notice that the days of the week are out of order so now we are going to order them
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Now we aggregate the ride time based on the average for the user type
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
```
## Analysis
Now we analyze data based on type and weekday
```{r}
#install.packages("lubridate")
library(lubridate)
?lubridate::date
```

```{r}
library(dplyr)
library(ggplot2)
all_trips_v2 %>% 
  mutate(weekday = lubridate::wday(start_time, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(usertype, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(usertype, weekday)					
```

## Visualizations
Let's visualize the number of rides by rider type
```{r}
all_trips_v2 %>% 
  mutate(weekday = lubridate::wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")
```

Let's create a visualization for average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = lubridate::wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge")
```

